---
title: "Adding a data source"
author: "ONgeoR contributors"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Adding a data source}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

Adding a source currently means changing the package. There is no exported
`add_source_template()` or `validate_source_registry()` workflow. The existing
retrieval helper supports Ontario LIO ArcGIS REST layers that return GeoJSON;
do not describe another provider or file format as supported until retrieval,
validation, and tests for it exist.

## Vet the source first

Before editing code, verify and record:

- the authoritative owner and publishing jurisdiction;
- the licence and whether redistribution or runtime retrieval is permitted;
- a stable ArcGIS REST MapServer layer URL and its `service_layer` path;
- geometry type, schema, key identifier and name fields;
- observed feature count and whether it exceeds one response page;
- whether generalized geometry is valid for this particular layer.

Feature counts and schemas can change. Treat the registry values as reviewed
metadata, not runtime guarantees.

## Add the registry entry

Add an entry under `sources:` in `inst/extdata/sources.yaml`. Existing entries
use exactly these fields:

```yaml
sources:
  example_source_id:
    name: "Human-readable authoritative layer name"
    service_layer: "LIO_OpenXX/0"
    geography_type: "boundary"
    feature_count: 123
    update_frequency: "unknown"
    key_fields: ["DOMAIN_ID", "ENGLISH_NAME"]
    license: "Open Government Licence - Ontario"
    source_url: "https://.../MapServer/0"
```

Use a stable snake-case source id. Confirm that `key_fields` exist in live
output and that the licence text and URL match the publisher. The package's
current YAML file contains a Unicode dash in existing licence values, but new
text can use an ASCII hyphen without changing meaning.

## Add a retrieval wrapper

Add a documented, exported wrapper in `R/retrieve.R`. It should state the
specific layer, parameters, return value, provenance, and any limitations
confirmed by live testing. The current LIO pattern is:

```{r wrapper, eval=FALSE}
#' Retrieve example boundaries
#'
#' @param simplify Logical. Whether to request generalized geometry.
#' @param refresh Logical. Whether to bypass the cache.
#' @return An `sf` object with provenance attributes.
#' @export
retrieve_example <- function(simplify = FALSE, refresh = FALSE) {
  fetch_lio_sf(
    service_layer = "LIO_OpenXX/0",
    source_name = "Human-readable authoritative layer name",
    simplify = simplify,
    refresh = refresh
  )
}
```

Choose `simplify` from observed geometry quality, not feature count alone.
Set `paginate = TRUE` only after confirming offset-based pagination works for
the layer. The cache key includes source name, layer, filter, simplification,
page size, and pagination mode, so changing any of those produces a distinct
cache entry. `refresh = TRUE` bypasses and replaces the matching cached entry.

If a wrapper accepts a filter, validate or constrain it before interpolating a
LIO `where` expression. Keep source-specific policy in the wrapper and common
HTTP, conversion, provenance, and cache behavior in `fetch_lio_sf()`.

Paginated retrieval requests 2,000 features per page and stops after 20 pages,
so the safety ceiling is 40,000 features. Retrieved row counts are checked
against the registry and warn when they differ by more than 20%. The optional
`update_frequency` registry field records source-maintenance metadata; it is
currently informational only.

## Wire the CLI dispatch

The registry does not automatically make a source retrievable by the CLI.
Update both switches in `R/cli.R`:

1. `retrieve_by_source_id()` maps the source id to the wrapper and passes
   `refresh`.
2. `source_retrieve_call()` emits the wrapper call used in generated
   `reproduce.R` scripts.

For example:

```{r dispatch, eval=FALSE}
# In retrieve_by_source_id()
example_source_id = retrieve_example(refresh = refresh),

# In source_retrieve_call()
example_source_id = "retrieve_example()",
```

These functions are internal, but both paths need tests because one executes
retrieval and the other generates user-facing reproducibility code.

## Test without depending on the service

Use mocked HTTP responses or synthetic `sf` data for routine tests. Cover:

- wrapper arguments passed to `fetch_lio_sf()`;
- returned geometry and `source_url`, `source_name`, `retrieved_at` attributes;
- cache reuse and `refresh = TRUE` behavior;
- pagination termination, page combination, and the existing hard page cap
  when pagination is enabled;
- simplified and unsimplified query construction where supported;
- both CLI dispatch functions and unknown-id errors;
- identifier and name detection when the source will be used by `resolve()` or
  `build_crosswalk()`.

Live validation is still needed before accepting the source, but it should be
an explicit manual check rather than a network dependency in the test suite or
vignettes. Record observed feature count, schema, geometry type, paging, and
simplification results in the change description.

## Document and check

Run roxygen to update `NAMESPACE` and the generated `.Rd` file, then run the
focused tests, the full test suite, and `devtools::check()` with vignettes.
Review the generated reproducer call, confirm no live API request occurred
during package checking, run `git diff --check`, and inspect the diff for
credentials or downloaded data before proposing a commit.
