---
title: "Building geographic crosswalks"
author: "ONgeoR contributors"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Building geographic crosswalks}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

A crosswalk records how features in one polygon layer relate to features in
another. `build_crosswalk()` standardizes the result and includes provenance
needed to audit it later. This vignette uses synthetic polygons and requires no
network connection.

## A reproducible polygon example

```{r layers}
library(ONgeoR)

square <- function(xmin, ymin, xmax, ymax) {
  sf::st_polygon(list(rbind(
    c(xmin, ymin), c(xmax, ymin), c(xmax, ymax),
    c(xmin, ymax), c(xmin, ymin)
  )))
}

phu <- sf::st_sf(
  PHU_ID = c("P1", "P2"),
  PHU_NAME_ENG = c("West PHU", "East PHU"),
  geometry = sf::st_sfc(
    square(0, 0, 1, 1), square(1, 0, 2, 1), crs = 3857
  )
)
municipal <- sf::st_sf(
  MUNID = c("M1", "M2"),
  MUNICIPAL_NAME = c("West municipality", "East municipality"),
  geometry = sf::st_sfc(
    square(0.1, 0.1, 0.9, 0.9), square(1.1, 0.1, 1.9, 0.9), crs = 3857
  )
)

attr(phu, "source_name") <- "Synthetic PHU boundaries"
attr(phu, "source_url") <- "https://example.invalid/phu"
attr(phu, "retrieved_at") <- as.POSIXct("2026-01-01", tz = "UTC")
attr(municipal, "source_name") <- "Synthetic municipal boundaries"
attr(municipal, "source_url") <- "https://example.invalid/municipal"

crosswalk <- build_crosswalk(municipal, phu, method = "within")
crosswalk
```

`within` is appropriate when every source polygon should be wholly contained
by one target polygon and both layers use compatible, sufficiently precise
boundaries. It avoids counting polygons that merely touch.

Real municipal layers are simplified by default. Generalizing each layer can
move nominally shared borders enough that a municipal polygon is no longer
strictly within a PHU polygon. For those layers, `intersects` is often the
practical predicate. It can return more than one row for a feature that crosses
or touches multiple targets, so inspect duplicates rather than assuming a
one-to-one result.

## Inspect the audit fields

The canonical output contains source and target identifiers and names, the
registry-style source labels, the matching method, reserved distance field,
source URLs, and the target retrieval time.

```{r inspect}
names(crosswalk)
crosswalk[, c(
  "from_id", "from_name", "to_id", "to_name", "match_method",
  "source_url_from", "source_url_to", "retrieved_at"
)]
```

`match_distance_km` is currently `NA` because `build_crosswalk()` implements
topological polygon matching, not nearest-neighbour matching.

Export only after checking unmatched or duplicated source identifiers and the
provenance fields. This write is deliberately not run during the vignette
build:

```{r export, eval=FALSE}
write.csv(crosswalk, "municipal-to-phu.csv", row.names = FALSE)
```

## Live Ontario example

The following example requires network access or populated ONgeoR caches and
is not executed while this vignette builds.

```{r live-crosswalk, eval=FALSE}
municipal <- retrieve_municipal("upper")
phu <- retrieve_phu()

municipal_to_phu <- build_crosswalk(
  municipal,
  phu,
  method = "intersects"
)

map_layers(Municipalities = municipal, PHUs = phu)
```
