---
title: "Getting started with ONgeoR"
author: "ONgeoR contributors"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with ONgeoR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

ONgeoR helps a public-health analyst connect locations and boundaries to
Ontario health geography while retaining source provenance. This vignette uses
small synthetic layers so that it builds without a network connection. The
equivalent live calls are shown separately and are not run.

## Discover and retrieve sources

The package registry describes the available authoritative layers.

```{r registry}
library(ONgeoR)

head(list_sources())
get_source("phu_boundaries")
```

Retrieval functions download an `sf` layer, attach `source_url`, `source_name`,
and `retrieved_at` attributes, and save it in the ONgeoR cache. A normal call
reuses the cached result. Use `refresh = TRUE` when you deliberately want a new
copy from the source. Record the provenance attributes alongside exported
analysis results because the upstream layer can change.

The following calls require network access or an existing cache and therefore
are not executed while this vignette builds.

```{r live-retrieve, eval=FALSE}
phu <- retrieve_phu()
hospitals <- retrieve_moh_service_locations(service_type = "Hospital")

# Deliberately bypass the cache and retrieve a new copy.
phu_fresh <- retrieve_phu(refresh = TRUE)

attr(phu, "source_url")
attr(phu, "source_name")
attr(phu, "retrieved_at")
```

## Link locations to boundaries

Here are two synthetic health-unit polygons, two analyst locations, and three
facilities. The column names resemble the registered LIO schemas so that the
same code structure transfers to live data.

```{r synthetic-data}
square <- function(xmin, ymin, xmax, ymax) {
  sf::st_polygon(list(rbind(
    c(xmin, ymin), c(xmax, ymin), c(xmax, ymax),
    c(xmin, ymax), c(xmin, ymin)
  )))
}

phu_demo <- sf::st_sf(
  PHU_ID = c("A", "B"),
  PHU_NAME_ENG = c("West Demo PHU", "East Demo PHU"),
  geometry = sf::st_sfc(
    square(-80.0, 43.0, -79.5, 43.5),
    square(-79.5, 43.0, -79.0, 43.5),
    crs = 4326
  )
)
attr(phu_demo, "source_name") <- "Synthetic PHU boundaries"
attr(phu_demo, "source_url") <- "https://example.invalid/phu"
attr(phu_demo, "retrieved_at") <- as.POSIXct("2026-01-01", tz = "UTC")

locations <- data.frame(
  location = c("Clinic request 1", "Clinic request 2"),
  lon = c(-79.75, -79.25),
  lat = c(43.25, 43.25)
)

linked <- link(locations, phu_demo)
linked[, c("location", "PHU_ID", "PHU_NAME_ENG", "target_url")]
```

For live Ontario boundaries, the corresponding operation is:

```{r live-link, eval=FALSE}
phu <- retrieve_phu()
linked <- link(locations, phu)
```

## Find or resolve facilities

`nearest()` performs a spatial proximity search. `resolve()` instead looks up
attributes by an identifier or name; it is not a spatial operation.

```{r nearest-resolve}
facilities_demo <- sf::st_as_sf(
  data.frame(
    MOH_SERVICE_PROVIDER_IDENT = c("F001", "F002", "F003"),
    ENGLISH_NAME = c("West Hospital", "Central Clinic", "East Hospital"),
    lon = c(-79.78, -79.52, -79.22),
    lat = c(43.24, 43.27, 43.24)
  ),
  coords = c("lon", "lat"), crs = 4326
)
attr(facilities_demo, "source_name") <- "Synthetic facilities"
attr(facilities_demo, "source_url") <- "https://example.invalid/facilities"
attr(facilities_demo, "retrieved_at") <- as.POSIXct("2026-01-01", tz = "UTC")

nearby <- nearest(locations, facilities_demo, k = 2)
nearby[, c("location", "rank", "ENGLISH_NAME", "distance_km")]

resolve(facilities_demo, "F002")
resolve(facilities_demo, "hospital", by = "name")
```

The live versions use the retrieved facility layer:

```{r live-nearest-resolve, eval=FALSE}
hospitals <- retrieve_moh_service_locations(service_type = "Hospital")
nearby <- nearest(locations, hospitals, k = 3, max_dist_km = 25)
facility <- resolve(hospitals, "12345")
named_facilities <- resolve(hospitals, "general", by = "name")
```

## Audit and map the result

Returned tables carry provenance columns such as `source_url`, `target_url`,
and `retrieved_at`. Inspect them before exporting, particularly after a cache
refresh. Interactive maps are useful for checking surprising assignments and
nearest matches.

```{r live-map, eval=FALSE}
map_layers(PHUs = phu, Hospitals = hospitals)
map_nearest(locations, hospitals, k = 3, max_dist_km = 25)
```

These maps are review aids, not substitutes for checking source metadata,
geometry precision, unmatched records, and the provenance columns in the
analysis output.
