---
title: "Launching and using the ONgeoR Shiny app"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Launching and using the ONgeoR Shiny app}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

ONgeoR ships a browser-based Shiny application that lets you build spatial
crosswalks and run nearest-match searches without writing any R code. This
vignette describes how to launch the app and use its two main tabs.

## Launching the app

Open a browser-based session with a single call:

```{r launch, eval=FALSE}
ONgeoR::run_app()
```

`run_app()` calls `shiny::runApp()` on the bundled app directory
(`inst/shiny`). By default Shiny opens a browser tab automatically. You can
pass additional arguments to control the port or host:

```{r launch-port, eval=FALSE}
ONgeoR::run_app(port = 4321, launch.browser = FALSE)
```

The app requires the `shiny` and `bslib` packages; `run_app()` checks for them
at startup and prompts you to install them if they are missing.

## The two tabs

### Link tab

The **Link** tab retrieves two registered Ontario GeoHub or LIO layers and
links them spatially.

Workflow:

1. **Base layer** -- Select a source from the "Base layer" dropdown. This is
   the layer whose features will be matched (e.g. Public Health Unit
   boundaries).
2. **Overlay source** -- Select a second source from the "Overlay source"
   dropdown (e.g. MOH Service Locations).
3. **Preview** -- Click **Preview** to retrieve both layers and draw them on
   an interactive Leaflet map. Inspect the map to confirm the layers align as
   expected before linking.
4. **Link** -- Click **Link** to run `link()` against the two layers. The
   result table appears below the map.

### Find Nearest tab

The **Find Nearest** tab performs a nearest-match search and draws connector
lines between each source point and its nearest target.

Workflow:

1. **Source points** -- Upload a CSV of your own starting points using the
   "Source points (CSV with lon/lat columns)" file input (e.g. a set of
   clinic locations).
2. **Target source** -- Select the registered candidate layer to search
   within (e.g. MOH Service Locations).
3. **k** -- Set the number of nearest targets to return per source point
   (default 1).
4. **max_dist_km** -- Cap results to targets within this radius.
5. **Preview on map / Find nearest** -- Preview draws the uploaded points
   and target layer; Find nearest runs the search. The map shows connector
   lines from each source to its nearest target(s), and the result table
   lists the distances.

## Download buttons

Each tab has its own download buttons, which become available after the
corresponding operation completes.

Link tab:

| Button | Output |
|---|---|
| **Download map** | `map.html` -- the Leaflet map as a standalone HTML file |
| **Download results** | `mapping.csv` (or `linked.csv` for raster linking) |
| **Download script** | `reproduce.R` -- an R script that reproduces the result from scratch |

Find Nearest tab:

| Button | Output |
|---|---|
| **Download map** | `map.html` -- the Leaflet map as a standalone HTML file |
| **Download results** | `nearest.csv` -- the nearest-match table |

The `map.html` file embeds the widget's JavaScript and data, so it opens
without R installed; note the basemap tiles are still fetched from the tile
provider, so viewing it requires network access.

## A note on automated testing

The source picker dropdowns use selectize.js, a JavaScript-enhanced widget
that wraps a standard HTML select element. When writing automated tests with
`shinytest2`, the underlying input value can be set programmatically:

```r
# Works: sets the underlying Shiny input value
app$set_inputs(base_layer = "phu_boundaries")
```

However, the visible selectize dropdown does not update its displayed text
until the JavaScript layer processes the change. Tests that rely on the widget
visually showing the selected value must follow the `set_inputs()` call with
either `app$click()` on a trigger element or direct JavaScript injection via
`app$execute_script()`.

Because of this constraint, the ONgeoR test suite covers the **server-side
logic** (retrieval, linking, nearest-match computation, and download
generation) in isolation rather than driving the full UI flow through a
headless browser. This is intentional: the spatial and data correctness
properties are best tested at the function level, while UI integration tests
would add fragility without exercising new logic paths.

If you want to add UI-level coverage, `shinytest2::AppDriver` is the
recommended approach. See `vignette("shinytest2", package = "shinytest2")` for
details on recording and replaying interactions with selectize widgets.
